Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-07 _tex/index.tex
Built site for gh-pages

However, the structural approach offers an additional key advantage over
the potential outcomes approach: it enables the graphical representation
of any system through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs serve as heuristics,
effectively conveying the assumed causal structure of a system. A
heuristic is a strategy that simplifies information to make decisions
more quickly, efficiently, and sometimes more accurately than complex
methods \citep{Chow_2015}. Consequently, DAGs do not represent detailed
statistical models but allow researchers to deduce which statistical
models can provide valid causal inferences, assuming the causal
structure depicted in the DAGs are accurate \citep{McElreath_2020}.

@@

In statistics, one application incorporating concepts from graph theory
is causal inference. Specifically, the structural approach to causal
inference uses directed acyclic graphs (DAG) to provide a graphical and
formal representation of the causal structure of a system
\citep{Neal_2020}. In this context, a \emph{graph} denotes a collection
of nodes connected by edges, where nodes represent random variables. The
term \emph{directed} indicates the edges of the graph extend from one
node to another, with arrows showing the direction of causal influence.
Moreover, the term \emph{acyclic} indicates the causal influences do not
form a loop, meaning the influences do not cycle back on themselves
\citep{McElreath_2020}.

DAGs offer two key advantages for modeling causality. Firstly, DAGs
represent causal relations in a nonparametric and fully interactive
manner. This feature implies that feasible causal analysis strategies
usually do not require specifying the type of data or the nature of the
functional dependence among variables \citep{Morgan_et_al_2014}.
Secondly, regardless of complexity, DAGs can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. These building blocks can be depicted
in three ways: the magnified representation, the standard
representation, and the structural causal model form
\citep[SCM,][]{Morgan_et_al_2014}.

Often, the DAGs omit the exogenous variables for simplicity, resulting
in the \emph{standard} representation. However, including exogenous
variables in a graph can be beneficial in some scenarios, as their
presence can reveal potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}, concepts explored in the
following section. The standard representation is illustrated in the
middle panels of Figure~\ref{fig-dags_scms}.

@@

Lastly, the right panels of Figure~\ref{fig-dags_scms} depict the SCM
form of the five fundamental building blocks. SCMs are formal
mathematical models that represent causal relationships within a system
or population \citep{Hitchcock_2024}. An SCM is defined by a set of
endogenous variables \(V\), a set of exogenous variables \(E\), and a
set of functions \(F=\{f_{X},f_{Z},f_{Y}\}\)
\citep{Pearl_2009, Neal_2020}. These functions, referred to as
structural equations, specify each endogenous variable as nonparametric
functions of other variables. Moreover, SCMs use the symbol `\(:=\)' to
indicate the variables' asymmetrical causal dependence and the symbol
`\(\perp\!\!\!\perp\)' to represent \emph{d-separation}, which roughly
equates to the concept of variable independence. The concepts of
d-separation and causal (in)dependence are explored in the following
section.

@@

A careful examination of Figure~\ref{fig-dags_scms} highlights the
assumptions underlying these building blocks. Figures
\ref{fig-mdag_bb1}, \ref{fig-sdag_bb1}, and SCM \ref{fig-scm_bb1} depict
two unconnected nodes, representing a scenario where variables \(X\) and
\(Y\) are not causally related. Figures \ref{fig-mdag_bb2},
\ref{fig-sdag_bb2}, and SCM \ref{fig-scm_bb2} illustrate two connected
nodes, showing a scenario where a \emph{parent} node \(X\) exerts a
causal influence on a \emph{child} node \(Y\). Consequently, \(Y\) is
considered a \emph{descendant} of \(X\). Figures \ref{fig-mdag_bb3},
\ref{fig-sdag_bb3}, and SCM \ref{fig-scm_bb3} depict a \emph{chain} or
\emph{pipe}, where \(X\) influences \(Z\), and \(Z\) influences \(Y\).
In this configuration, \(X\) is a parent node of \(Z\), and \(Z\) is a
parent node of \(Y\). Furthermore, the DAGs show that \(X\) is an
\emph{ancestor} of \(Y\) and that \(Z\) fully \emph{mediates} the
relationship between these variables. Figures \ref{fig-mdag_bb4},
\ref{fig-sdag_bb4}, and SCM \ref{fig-scm_bb4} illustrate a \emph{fork},
where variables \(X\) and \(Y\) are both influenced by \(Z\). Here,
\(Z\) is a parent node of \(X\) and \(Y\). Finally, Figures
\ref{fig-mdag_bb5}, \ref{fig-sdag_bb5}, SCM \ref{fig-scm_bb5} depict a
\emph{collider}, also known as \emph{inmorality}, where variables \(X\)
and \(Y\) are concurrent causes of \(Z\). In this configuration, \(X\)
and \(Y\) are not causally related to each other but both influence
\(Z\). Additionally, in all SCMs, the errors are assumed to be mutually
independent of each other and of all other variables in the graph, as
evidenced by the pairwise relations \(e_{X} \perp\!\!\!\perp e_{Y}\),
\(e_{X} \perp\!\!\!\perp e_{Z}\), and \(e_{Z} \perp\!\!\!\perp e_{Y}\).

@@

\includegraphics[width=1\textwidth,height=\textheight]{images/figures/ACflow.png}

\subsection{But where does it all fit?}\label{sec-background-where}

>>>>

However, the structural approach offers an additional key advantage over
the potential outcomes approach: it enables the graphical representation
of any system through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs serve as heuristics,
effectively conveying the assumed causal structure of a system
\citep{Pearl_et_al_2016}. A heuristic is a strategy that simplifies
information to make decisions more quickly, efficiently, and sometimes
more accurately than complex methods \citep{Chow_2015}. Consequently,
DAGs do not represent detailed statistical models but allow researchers
to deduce which statistical models can provide valid causal inferences,
assuming the causal structure depicted in the DAGs are accurate
\citep{McElreath_2020}.

@@

In statistics, one application incorporating concepts from graph theory
is causal inference. Specifically, the structural approach to causal
inference uses directed acyclic graphs (DAG) to provide a graphical and
formal representation of the causal structure of a system
\citep{Neal_2020}. In this context, a \emph{graph} denotes a collection
of nodes connected by edges, where nodes represent random variables. The
term \emph{directed} indicates the edges of the graph extend from one
node to another, with arrows showing the direction of causal influence.
Moreover, the term \emph{acyclic} indicates the causal influences do not
form a loop, meaning the influences do not cycle back on themselves
\citep{McElreath_2020}. Several examples of DAGs can be found in
Figure~\ref{fig-dags_scms}.

DAGs offer two key advantages for modeling causal structures. Firstly,
they represent causal relations in a nonparametric and fully interactive
manner. This feature allows for feasible causal analysis strategies
without needing the specification of the type of data or the nature of
the functional dependence among variables \citep{Morgan_et_al_2014}.
Secondly, regardless of complexity, DAGs can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature enables the
decomposition of complex structures into basic building blocks,
facilitating the analysis of these structures by focusing on the causal
assumptions associated with each building block individually
\citep{McElreath_2020}. These building blocks can be represented in
three ways: the magnified representation, the standard representation,
and the structural causal model form \citep[SCM,][]{Morgan_et_al_2014}.

@@

@@

Often, DAGs omit the exogenous variables for simplicity, resulting in
the \emph{standard} representation. However, including exogenous
variables in a graph can be beneficial in some scenarios, as their
presence can reveal potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}, concepts explored in the
following section. The standard representation is illustrated in the
middle panels of Figure~\ref{fig-dags_scms}.

Lastly, the right panels of Figure~\ref{fig-dags_scms} depict the SCM
form of the fundamental building blocks. SCMs are formal mathematical
models defined by a set of endogenous variables \(V\), a set of
exogenous variables \(E\), and a set of functions
\(F=\{f_{X},f_{Z},f_{Y}\}\) \citep{Pearl_2009, Neal_2020}. These
functions, referred to as structural equations, specify each endogenous
variable as nonparametric functions of other variables. Moreover, SCMs
use the symbol `\(:=\)' to indicate the variables' asymmetrical causal
dependence and the symbol `\(\perp\!\!\!\perp\)' to represent
\emph{d-separation}, which roughly equates to the concept of variable
independence. The concepts of d-separation and causal (in)dependence are
explored in the following section.

A careful examination of Figure~\ref{fig-dags_scms} highlights the
assumptions underlying these building blocks. Figures
\ref{fig-mdag_bb1}, \ref{fig-sdag_bb1}, and SCM \ref{fig-scm_bb1} depict
two unconnected nodes, representing a scenario where variables \(X\) and
\(Y\) are not causally related. Figures \ref{fig-mdag_bb2},
\ref{fig-sdag_bb2}, and SCM \ref{fig-scm_bb2} illustrate two connected
nodes, showing a scenario where a \emph{parent} node \(X\) exerts a
causal influence on a \emph{child} node \(Y\). Consequently, \(Y\) is
considered a \emph{descendant} of \(X\). Figures \ref{fig-mdag_bb3},
\ref{fig-sdag_bb3}, and SCM \ref{fig-scm_bb3} depict a \emph{chain} or
\emph{pipe}, where \(X\) influences \(Z\), and \(Z\) influences \(Y\).
In this configuration, \(X\) is a parent node of \(Z\), and \(Z\) is a
parent node of \(Y\). This creates a \emph{directed path} between \(X\)
and \(Y\). Consequently, \(X\) is an \emph{ancestor} of \(Y\), and \(Z\)
fully \emph{mediates} the relationship between the two. Figures
\ref{fig-mdag_bb4}, \ref{fig-sdag_bb4}, and SCM \ref{fig-scm_bb4}
illustrate a \emph{fork}, where variables \(X\) and \(Y\) are both
influenced by \(Z\). Here, \(Z\) is a parent node of \(X\) and \(Y\).
Finally, Figures \ref{fig-mdag_bb5}, \ref{fig-sdag_bb5}, SCM
\ref{fig-scm_bb5} depict a \emph{collider}, also known as
\emph{inmorality}, where variables \(X\) and \(Y\) are concurrent causes
of \(Z\). In this configuration, \(X\) and \(Y\) are not causally
related to each other but both influence \(Z\). Additionally, in all
SCMs, the errors are assumed to be mutually independent of each other
and of all other variables in the graph, as evidenced by the pairwise
relations \(e_{X} \perp\!\!\!\perp e_{Y}\),
\(e_{X} \perp\!\!\!\perp e_{Z}\), and \(e_{Z} \perp\!\!\!\perp e_{Y}\).

\includegraphics[width=0.9\textwidth,height=\textheight]{images/figures/ACflow.png}

\subsection{A motivating example}\label{sec-framework-example}

@@
